Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes in communities & finishing career path section

\subsection{Credit, Citation, Impact}

@@

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al.~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al.~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructure development built-in,
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

@@

Drawing on experiences from high-energy physics, \cite{Vay_WSSSPE} proposed
developing teams of technical specialists able to overcome a lack of
coordination between projects.  Maximization of scientific output requires
maximizing the usability of the scientific codes while minimizing the cost of
developing and supporting those codes.  This included targeting different
architectures for their software to be deployed, as well as coordination
between technically-focused individuals and usage of a common scripting
language between projects.  Instead of fragmenting the development of
simulation codes across institutions, the paper suggests that a cohesive
strategy reducing duplication and increasing coordination will broadly increase
the efficiency across institutions.  The approach proposed is of de-fragmenting
the existing ecosystem in a non-disruptive way.

@@

In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
source software development at Kitware.  In particular, they focus on their
success at growing their community of users through enabling multiple channels
of communication, directly reaching out to individuals, and lowering the
barrier to entry for contributions. This involves providing clear,
test-oriented and review-based mechanisms for evaluating contributions,
permissive licenses, and a service-based model for sustaining development.
This model enables Kitware to review both public funding, as well as private
funding to support improvements and targeted developments of the software.

@@

\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
collaborations supporting tool development for Open Science.  This software
collective, organized around the statistical programming environment R,
develops access mechanisms for data repositories and attempts to reduce the
barrier to entry for individuals wanting to access data repositories and study
the data contained therein.  The collective fosters direct collaboration
between individuals and data providers, designed to ``train academics in
reproducible science workflows focused around R.''  The two central challenges
to this goal were identified as engagement of existing users within ecology and
evolutionary biology (EEB), and how the community could make inroads and traction
in other disciplines. Currently, the collective is exploring addressing these
challenges through use of social media and holding workshops and hackathons.
This helps to both raise the profile of the collective within EEB and in other
domains.  However, the overarching challenge identified in the paper was that
of incentivizing maintenance of software, which is difficult in academia.

@@

In terms of possible solutions, the Developing and Supporting Software discussion (\S\ref{sec:devel}) said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketches of asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
However, it was also noted it can be very difficult to measure where people are using small bits of your code.

Open questions:

What are the possible career paths for a specialist software developer working as part of a scientific research group?

>>>>

\subsection{Credit, Citation, and Impact}

@@

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al.~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al.~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructure development built-in,
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

@@

Drawing on experiences from high-energy physics, Vay et al.~\cite{Vay_WSSSPE} proposed
developing teams of technical specialists able to overcome a lack of
coordination between projects.  Maximization of scientific output requires
maximizing the usability of the scientific codes while minimizing the cost of
developing and supporting those codes.  This included targeting different
architectures for their software to be deployed, as well as coordination
between technically-focused individuals and usage of a common scripting
language between projects.  Instead of fragmenting the development of
simulation codes across institutions, the paper suggests that a cohesive
strategy reducing duplication and increasing coordination will broadly increase
the efficiency across institutions.  The approach proposed is of de-fragmenting
the existing ecosystem in a non-disruptive way.

@@

Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of open
source software development at Kitware.  In particular, they focus on their
success at growing their community of users through enabling multiple channels
of communication, directly reaching out to individuals, and lowering the
barrier to entry for contributions. This involves providing clear,
test-oriented and review-based mechanisms for evaluating contributions,
permissive licenses, and a service-based model for sustaining development.
This model enables Kitware to review both public funding, as well as private
funding to support improvements and targeted developments of the software.

@@

Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
collaborations supporting tool development for Open Science.  This software
collective, organized around the statistical programming environment R,
develops access mechanisms for data repositories and attempts to reduce the
barrier to entry for individuals wanting to access data repositories and study
the data contained therein.  The collective fosters direct collaboration
between individuals and data providers, designed to ``train academics in
reproducible science workflows focused around R.''  The two central challenges
to this goal were identified as engagement of existing users within ecology and
evolutionary biology (EEB), and how the community could make inroads and traction
in other disciplines. Currently, the collective is exploring addressing these
challenges through use of social media and holding workshops and hackathons.
This helps to both raise the profile of the collective within EEB and in other
domains.  However, the overarching challenge identified in the paper was that
of incentivizing maintenance of software, which is difficult in academia.

@@

In the Policy discussion (\S\ref{sec:policy}), career paths were not discussed explicitly, though the discussion about credit, citation, and impact focused on how those who create or contribute to software are recognized; this recognition (or the current lack thereof) is tied to career paths (or their lack) for software developers.

The Communities discussion (\S\ref{sec:community}) did not directly address career paths, though one of the papers~\cite{Loffler_WSSSPE} that was part of this section also makes a similar policy point, in a section about career paths. Specifically, it says ``The most severe problem for developers in most computational sciences currently is that while most of the work is done creating hopefully well-written, sustainable software, the academic success is often exclusively tied to the solution of the scientific problem the software was designed for. Tasks that from a software engineering standpoint are essential, e.g., high usability, well-written and updated documentation, or porting infrastructure to new platforms, are not rewarded within this system.''

In terms of possible solutions, the Developing and Supporting Software discussion said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketch requirement: asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
However, it was also noted it can be very difficult to measure where people are using small bits of your code.  And the possible career paths for a specialist software developer working as part of a scientific research group in a university are certainly not yet clear.
